Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-11 bbbq_article.tex
Comments are visible in docx

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
Because the peptide-binding grooves of most haplotypes 
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
peptide fragments derived from transmembrane helices (TMHs) 
are predicted to be presented more often than expected 
based on their abundance in the proteome.
We show that the over-presentation of TMH-derived peptides is likely general, 
as it is predicted for diverse microbiota and viruses 
and for both MHC-I and MHC-II
However, the physiological reason of why membrane proteins might be 
over-presented is unclear
In this study, we show the TMHs are evolutionarily more conserved, 
as relatively less single nucleotide polymorphisms (SNPs) 
are present in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytosolic protein regions. 
Thus, our findings suggest that T cells might respond 
more to membrane proteins, because these are evolutionary more conserved.
We speculate that TMHs therefor might be less prone to escape mutations 
that enable pathogens to evade T cell responses.

@@

% GvdB: Not really relevant. You don’t really need to explain the role of DCs. 
% What is far more important is the concept of what an MHC is: 
% a protein complex that contains a fold, called the peptide-binding groove, 
% which presents peptides. Also immediately explain that MHC-I 
% presents peptides of about 9-amino acids in length to cytolytic T cells, 
% whereas MHC-II presents longer peptides to helper T cells.

@@

Any individual's immune system detects only a fraction of all possible
peptide fragments.
For humans, the MHC proteins are encoded by the
HLA (Human Leukocyte Antigens) genes.
There are three genes encoding for MHC-I, which are HLA-A, HLA-B and HLA-C,
for MHC-II there are three major genes, which are HLA-DR, HLA-DQ, HLA-DP.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition})
\frans{Is dit zo?} 
\richel{Volgens de (weggegevallen) referentie naar Lund et al., 2004, 'Definition of supertypes for {HLA} molecules using clustering of specificity matrices' wel}.
The HLA region of humans is highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different HLA gene is called 
an MHC haplotype (\cite{marsh2010nomenclature}).

@@

However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
We hypothesized is that the presentation of 
hydrophobic/TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to these requirements, 
the amino acids genuinely present in TMH 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
\frans{Analyzing the occurrence of single nucleotide polymorphisms that were predicted to be inside ...}

@@

The first query was a call to the 'gene' database for the 
term 'membrane protein' (in all fields) for the organism \textbf{Homo sapiens}.
This resulted in 1077 gene IDs
\richel{
  This was December 2020.
  At 2021-03-01, one will get 1130 gene IDs.
  Also, one of the gene IDs that was valid back then,
  has been obsoleted.
}.
The next query was a call to the 'gene' database 
to obtain the gene names from the gene IDs.
Per gene name, the 'dbSNP' NCBI database is queried for 
variations associated with the gene name.

@@

\richel{I suggest to move this section to the Supplementary Materials}

@@

\geert{A narrative was missing a bit. Why was this done?}
To confirm that peptides stemming from TMHs were presented in MHC-I and MHC-II,
we reanalyzed published peptide elution studies.
Table \ref{tab:elution} shows the percentage of epitopes derived 
from a TMH
found in MHC-I and MHC-II elution 
studies (\cite{schellens2015comprehensive} and \cite{bergseng2015different} respectively), 
for the two topology prediction tools TMHMM and 
PureseqTM (\cite{krogh2001predicting} and \cite{wang2019efficient} respectively). 
Regardless of the prediction, 
at least 100 epitopes were predicted to be derived from a TMH. 
\geert{Also add a conclusion at the end of the paragraph.}
From these findings, we conclude that PureseqTM and TMHMM 
result in prediction of a similar number of TMHs 
and that peptides derived from TMHs have been identified for both MHC-I and -II.

@@

The actually number of SNPs found in TMHs
was $n_{SNP\_TMH} = 3,903$, which is lower than the expectation
of $E(n_{SNP\_TMH}) \approx 4056$.
To test whether this difference is significant,
we used a binomial Poisson distribution, where the number of trails,
$n$ equals the number of SNPs, which is 21,738. The probability of success
for TMP $i$, $p\_i$, is the percentage of each TMP that is TMH.
We calculated the probability of having 3,903 or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals \richel{recalculate}
% $1.6201 \cdot 10^{-9}$.

@@

\richel{How many SNPs in TMHs?}
See figure \ref{fig:f_tmh_ncbi} to see which percentages
are TMH of the transmembrane proteins.

@@

We decided to split this analysis based on the number of TMHs
a protein has \richel{Why? Provide brief rationale again. Thus TMH-TMH interactions or something}. We hypothesize that single-spanners (i.e. proteins
with one TMH predicted) are less conserved, when compared to multi-spanners,
as single spanners just need to span a membrane, while multi-spanners
might have interactions between their TMHs, 
for example to accommodate active sites, and 
thus might have additional structural constraints.
As can be seen in figure \ref{fig:f_snps_found_and_expected_per_spanner}, 
both single- and multi-spanners are evolutionarily conserved.

@@

It is known that an individual's haplotype
influences the dynamics of a pathogen (see,
for example, \cite{eccleston2017host} for HIV) 
\geert{Too vague. How?}.
The bioinformatics pipeline developed in this
study may be interesting to make in silico
predictions of a SARS-CoV-2 infection.
The most prominent example is the (MHC-I)
haplotype HLA-A*24:02,
which is predicted to present four times as much epitopes derived 
from SARS-CoV-2 TMHs than expected by chance.
This may help the immune system to detect a SARS-CoV-2 infection earlier,
and change the infection's dynamics.

@@

\geert{
  Discussion section needs some work, 
  as here interpretation of the results needs to be given. 
  What is the consequence of the findings? Why is it important.

@@

  So offer interpretations. 
  The second part of the introduction would fit here well. I can also help with this.
}

@@

We assume that epitopes 
derived from TMHs are actually presented by MHC-I and MHC-II.
Our analysis (at subsection \ref{subsec:elution_studies}) 
finds that 1.4\% (for MHC-I) and 4.0% (for MHC-II) 
of the epitopes presented 
overlap with a TMH.
It is unsure if this a true finding or a false positive.
One source of noise would stem from imperfect predictions
by the topology prediction software used. 
We have counteracted this by using two different topology prediction
software and confirm these have similar results.
Two biological sources of noise \geert{What do you mean by noise?}
are defective ribosomal products (DRiPs) (\cite{yewdell1996defective}) 
and peptide fusion \cite{delong2016pathogenic},
two processes that both result in peptide fragments
that did not originate from genome directly.
Note however, that both elution studies have a
methodological bias caused by mass spectronomy,
as hydrophobic fragments are detected less 
by mass spectronomy.

@@

\geert{This might be relevant, but you need to explain why it is relevant. Thus compare your findings with previous efforts to look at conservation of TMHs}
% \paragraph{synonymous mutations}
%
In our evolutionary experiment, 
we removed variations that were synonymous mutations (i.e.
resulted in the same amino acid, from a different genetic code) 
from our analysis.
There is evidence, however, that these synonymous mutations
do have an effect and may even be evolutionary selected 
for (\cite{hunt2009silent}).
As the possible effect of synonymous mutations is ignored by our
topology prediction software, we do so as well.

@@

We thank the Center for Information Technology of the University 
of Groningen for its support and for providing access to the Peregrine 
high performance computing cluster. 
Additionally, we would like to thank Sci-Hub (\cite{himmelstein2018sci})
for allowing us to read paywalled articles while working from home.
\geert{Should we include Maxim? He did the alignment of Mycobacterium tuberculosis. Let’s also ask Frans.}

>>>>

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
Because the peptide-binding grooves of most haplotypes 
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
peptide fragments derived from transmembrane helices (TMHs) 
are predicted to be presented more often than expected,
based on their abundance in the proteome.
We show that the over-presentation of TMH-derived peptides is likely general, 
as it is predicted for diverse microbiota and viruses 
and for both MHC-I and MHC-II
However, the physiological reason of why membrane proteins might be 
over-presented is unclear
In this study, we show the TMHs are evolutionarily more conserved, 
as relatively less single nucleotide polymorphisms (SNPs) 
are present in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytosolic protein regions. 
Thus, our findings suggest that T cells might respond 
more to membrane proteins, because these are evolutionary more conserved.
We speculate that TMHs therefor might be less prone to escape mutations 
that enable pathogens to evade T cell responses.

@@

@@

Any individual's immune system detects only a fraction of all possible
peptide fragments.
For humans, the MHC proteins are encoded by the
HLA (Human Leukocyte Antigens) genes.
There are three genes encoding for MHC-I, which are HLA-A, HLA-B and HLA-C,
for MHC-II there are three major genes, which are HLA-DR, HLA-DQ, HLA-DP.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition})
[FB: is that really so?]
[RB: yes, according to the removed reference to Lund et al., 2004, 'Definition of supertypes for {HLA} molecules using clustering of specificity matrices'] 
The HLA region of humans is highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different HLA gene is called 
an MHC haplotype (\cite{marsh2010nomenclature}).

@@

However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
We hypothesized is that the presentation of 
hydrophobic/TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to these requirements, 
the amino acids genuinely present in TMH 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
[FB: Analyzing the occurrence of single nucleotide polymorphisms 
that were predicted to be inside ...]

@@

The first query was a call to the 'gene' database for the 
term 'membrane protein' (in all fields) for the organism \textbf{Homo sapiens}.
This resulted in 1077 gene IDs
[RJCB:
  This was December 2020.
  At 2021-03-01, one will get 1130 gene IDs.
  Also, one of the gene IDs that was valid back then,
  has been obsoleted.
].
The next query was a call to the 'gene' database 
to obtain the gene names from the gene IDs.
Per gene name, the 'dbSNP' NCBI database is queried for 
variations associated with the gene name.

@@

[RJCB: I suggest to move this section to the Supplementary Materials]

@@

[GVDB: A narrative was missing a bit. Why was this done?]
To confirm that peptides stemming from TMHs were presented in MHC-I and MHC-II,
we reanalyzed published peptide elution studies.
Table \ref{tab:elution} shows the percentage of epitopes derived 
from a TMH
found in MHC-I and MHC-II elution 
studies (\cite{schellens2015comprehensive} and \cite{bergseng2015different} respectively), 
for the two topology prediction tools TMHMM and 
PureseqTM (\cite{krogh2001predicting} and \cite{wang2019efficient} respectively). 
Regardless of the prediction, 
at least 100 epitopes were predicted to be derived from a TMH. 
[GVDB: Also add a conclusion at the end of the paragraph.]
From these findings, we conclude that PureseqTM and TMHMM 
result in prediction of a similar number of TMHs 
and that peptides derived from TMHs have been identified for both MHC-I and -II.

@@

The actually number of SNPs found in TMHs
was $n_{SNP\_TMH} = 3,903$, which is lower than the expectation
of $E(n_{SNP\_TMH}) \approx 4056$.
To test whether this difference is significant,
we used a binomial Poisson distribution, where the number of trails,
$n$ equals the number of SNPs, which is 21,738. The probability of success
for TMP $i$, $p\_i$, is the percentage of each TMP that is TMH.
We calculated the probability of having 3,903 or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals [RJCB: recalculate]
$1.6201 \cdot 10^{-9}$.

@@

[RJCB: How many SNPs in TMHs?]
See figure \ref{fig:f_tmh_ncbi} to see which percentages
are TMH of the transmembrane proteins.

@@

We decided to split this analysis based on the number of TMHs
a protein has 
[GVDB: Why? Provide brief rationale again. 
Thus TMH-TMH interactions or something]. 
We hypothesize that single-spanners (i.e. proteins
with one TMH predicted) are less conserved, when compared to multi-spanners,
as single spanners just need to span a membrane, while multi-spanners
might have interactions between their TMHs, 
for example to accommodate active sites, and 
thus might have additional structural constraints.
As can be seen in figure \ref{fig:f_snps_found_and_expected_per_spanner}, 
both single- and multi-spanners are evolutionarily conserved.

@@

It is known that an individual's haplotype
influences the dynamics of a pathogen (see,
for example, \cite{eccleston2017host} for HIV) 
[GVDB: Too vague. How?].
The bioinformatics pipeline developed in this
study may be interesting to make in silico
predictions of a SARS-CoV-2 infection.
The most prominent example is the (MHC-I)
haplotype HLA-A*24:02,
which is predicted to present four times as much epitopes derived 
from SARS-CoV-2 TMHs than expected by chance.
This may help the immune system to detect a SARS-CoV-2 infection earlier,
and change the infection's dynamics.

@@

[GVDB: 
  Discussion section needs some work, 
  as here interpretation of the results needs to be given. 
  What is the consequence of the findings? Why is it important.

@@

  So offer interpretations. 
  The second part of the introduction would fit here well. I can also help with this.
]

@@

We assume that epitopes 
derived from TMHs are actually presented by MHC-I and MHC-II.
Our analysis (at subsection \ref{subsec:elution_studies}) 
finds that 1.4\% (for MHC-I) and 4.0% (for MHC-II) 
of the epitopes presented 
overlap with a TMH.
It is unsure if this a true finding or a false positive.
One source of noise would stem from imperfect predictions
by the topology prediction software used. 
We have counteracted this by using two different topology prediction
software and confirm these have similar results.
Two biological sources of noise [GVDB: What do you mean by noise?]
are defective ribosomal products (DRiPs) (\cite{yewdell1996defective}) 
and peptide fusion \cite{delong2016pathogenic},
two processes that both result in peptide fragments
that did not originate from genome directly.
Note however, that both elution studies have a
methodological bias caused by mass spectronomy,
as hydrophobic fragments are detected less 
by mass spectronomy.

@@

[GVDB: 
  This might be relevant, but you need to explain why it is relevant. 
  Thus compare your findings with previous efforts 
  to look at conservation of TMHs
]
% \paragraph{synonymous mutations}
%
In our evolutionary experiment, 
we removed variations that were synonymous mutations (i.e.
resulted in the same amino acid, from a different genetic code) 
from our analysis.
There is evidence, however, that these synonymous mutations
do have an effect and may even be evolutionary selected 
for (\cite{hunt2009silent}).
As the possible effect of synonymous mutations is ignored by our
topology prediction software, we do so as well.

@@

We thank the Center for Information Technology of the University 
of Groningen for its support and for providing access to the Peregrine 
high performance computing cluster. 
Additionally, we would like to thank Sci-Hub (\cite{himmelstein2018sci})
for allowing us to read paywalled articles while working from home.
[GVDB: 
  Should we include Maxim? 
  He did the alignment of Mycobacterium tuberculosis. 
  Let’s also ask Frans.
]
